Notes on Perturbational Quantities

### 2011-06-16 docs/pert_notes/pert.tex
more reorganization

Maestro solves an enthalpy equation.  
The full enthalpy equation is
\begin{eqnarray}
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h \Ub) + \frac{Dp_0}{Dt} 
+ \nabla\cdot \kth \nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext} \nonumber \\
&=& \underbrace{-\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub}_{-\nabla\cdot(\rho h\Ub)} 
+ \underbrace{\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r}}_{\frac{Dp_0}{Dt}} 
+ \nabla\cdot\kth\nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext}.
\end{eqnarray}
%The full temperature equation is
%\begin{equation}
%\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
%+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
%+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] - \sum_k\rho\xi_k\dot\omega_k 
%+ \rho H_{\rm nuc} + \rho H_{\rm ext}\right\}.
%\end{equation}
Due to Strang-splitting of the reactions,the call to {\tt
  react\_state} has already been made.  Hence, the goal is to compute
an edge state enthalpy starting from $(\rho h)^{(1)}$ using an
enthalpy equation that does not include the $\rho H_{\rm nuc}$ and
$\rho H_{\rm ext}$ terms, where were already accounted for in {\tt
  react\_state}, so our equation becomes
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub 
+ \psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla\cdot\kth\nabla T. \label{rhoh equation}
\end{equation}
We define the base state enthalpy evolution equation as
\begin{eqnarray}
\frac{\partial(\rho h)_0}{\partial t} &=& -\nabla\cdot[(\rho h)_0 w_0\eb_r] 
+ \frac{D_0p_0}{Dt} \nonumber \\
&=& -w_0\frac{\partial(\rho h)_0}{\partial r} 
- \underbrace{(\rho h)_0\frac{\partial w_0}{\partial r}+ \psi}_{"(\rho h)_0 ~ \text{force}"}
\enskip .\label{rhoh0 equation}
\end{eqnarray}
Subtracting (\ref{rhoh0 equation}) from (\ref{rhoh equation}) and
rearranging terms gives the perturbational enthalpy equation
\begin{eqnarray}
\frac{\partial(\rho h)'}{\partial t} &=& -\nabla\cdot[(\rho h)'\Ub] 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r} 
+ \nabla\cdot\kth\nabla T\nonumber \\
&=& -\Ub\cdot\nabla(\rho h)' \underbrace{- (\rho h)'\nabla\cdot\Ub 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r}
+ \nabla\cdot\kth\nabla T}_{"(\rho h)' ~ \text{force}"}, \label{rhohprime equation}
\end{eqnarray}

Alternately, we can consider an temperature evolution equation, derived
from enthalpy, as:
\begin{equation}
\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] 
+ \nabla \cdot \kth \nabla T
- \sum_k\rho\xi_k\omegadot_k 
+ \rho H_{\rm nuc} + \rho H_{\rm ext}\right\}.
\end{equation}
Again, we neglect the reaction terms, since that will be handled during
the reaction step, so we can write this as:
\begin{equation}
\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] 
+ \nabla \cdot \kth \nabla T \right \} \, .
\end{equation}

A final alternative is to consider an evolution equation for $h$
alone.  This can be derived by expanding the derivative of $(\rho h)$
in Eq.~\ref{rhoh equation} and subtracting off $h \times$ the
continuity equation (Eq.~\ref{rho equation}):
\begin{equation}
\frac{\partial h}{\partial t} = -\Ub \cdot \nabla h + \frac{1}{\rho}
\left \{ \psi + (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}
+ \nabla \cdot \kth \nabla T \right \} \, .
\end{equation}

@@

\begin{landscape}
\begin{table}[t]
\renewcommand{\arraystretch}{1.5}
{\small
\centering
\begin{tabular}{l|l|c|c|c|c}
\hline
\hline
{\tt species\_pred\_type} &   
{\tt enthalpy\_pred\_type} &
{quantity predicted} &
{intermediate} &
{species quantity} &
{final $(\rho h)$} \\[-5pt]
 &
 &
{in {\tt make\_edge\_scal}} &
{``enthalpy'' edge state} &
{available in {\tt mkflux}} &
{edge state} \\
\hline 
1 / {\tt predict\_rhoprime\_and\_X}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'$ & $X$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 2 / {\tt predict\_h} &
  $h$ & $h$ & $X$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'$ & $X$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h$ & $X$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
\hline
\hline
2 / {\tt predict\_rhoX}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'$ & $(\rho X)$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
2 / {\tt predict\_rhoX}  & 2 / {\tt predict\_h} &
  $h$ & $h$ & $(\rho X)$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
2 / {\tt predict\_rhoX}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'$ & $(\rho X)$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
2 / {\tt predict\_rhoX}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h$ & $(\rho X)$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
\hline
\hline
\end{tabular}
} % end \small
\end{table}
\end{landscape}

@@

\subsection{Computing $\rho h$ at edges}
We use an analogous procedure described in Section \ref{Computing rho
at edges} for computing $\rho$ at edges to compute $\rho h$ at edges.
However, instead of computing $(\rho h)_0$ to edges directly, we compute $\rho$
and $h$ separately, and multiply to get $(\rho h)_0$.

@@

\subsection{Summary of Forcing}

\begin{table*}[h]
\begin{center}
\caption{Forcing term into {\tt make\_edge\_scal} \newline}
\begin{tabular}{c|c}
\hline
\hline
{H PRED TYPE} &   {Force} \\
\hline \\[-3mm]
1  $((\rho h)^\prime)$ &  $(\Ubt \cdot \er) \frac{\partial p_0}{\partial r} - 
 (\rho h)^\prime \; \nabla \cdot (\Ubt+w_0 \er) - 
 \nabla \cdot (\Ubt (\rho h)_0) + \nabla \cdot \kth \nabla T$ \\[2 mm]
2  $(h)$ & $\frac{1}{\rho} \left [\psi + (\Ubt \cdot \er)
  \frac{\partial p_0}{\partial r} + \nabla \cdot \kth \nabla T \right ]$ \\[2 mm]
3  $(T)$ & $\frac{1}{\rho c_p} \left \{ (1 - \rho h_p) 
   \left [\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} \right ] \nabla \cdot \kth \nabla T \right \}$ \\[2 mm]
4  $(T)$ & $\frac{1}{\rho c_p} \left\{ (1 - \rho h_p) \left [\psi + (\Ubt \cdot \er)
\frac{\partial p_0}{\partial r}\right ] \nabla \cdot \kth \nabla T \right\}$ \\[2 mm]
\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[h]
\begin{center}
\caption{Quantity that goes into and out of {\tt make\_edge\_scal} \newline}
\begin{tabular}{c|c}
\hline
\hline
{h pred type} & {h/T } \\
\hline \\[-3mm]
1 & $(\rho h)^\prime$ \\
2 & $h$  \\
3 & $T$  \\
4 & $T$  \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

\begin{table*}
\begin{center}
\caption{Quantity that goes into {\tt mkflux} on edges \newline}
\begin{tabular}{c|c|c}
\hline
\hline
{h pred type} & {Species } & {h/T } \\
\hline
1 & $X$ and $\rho^\prime$ & $(\rho h)^\prime$ \\
2 & $X$ and $\rho^\prime$ & $h$ \\
3 & $X$ and $\rho^\prime$ & $(\rho h)^\prime$ \\
4 & $X$ and $\rho^\prime$ & $h$ \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

>>>>

Maestro solves an enthalpy equation.  
The full enthalpy equation is
\begin{eqnarray}
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h \Ub) + \frac{Dp_0}{Dt} 
+ \nabla\cdot \kth \nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext} \nonumber \\
&=& \underbrace{-\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub}_{-\nabla\cdot(\rho h\Ub)} 
+ \underbrace{\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r}}_{\frac{Dp_0}{Dt}} 
+ \nabla\cdot\kth\nabla T + \rho H_{\rm nuc} + \rho H_{\rm ext}.
\end{eqnarray}
%The full temperature equation is
%\begin{equation}
%\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
%+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
%+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] - \sum_k\rho\xi_k\dot\omega_k 
%+ \rho H_{\rm nuc} + \rho H_{\rm ext}\right\}.
%\end{equation}
Due to Strang-splitting of the reactions,the call to {\tt
  react\_state} has already been made.  Hence, the goal is to compute
an edge state enthalpy starting from $(\rho h)^{(1)}$ using an
enthalpy equation that does not include the $\rho H_{\rm nuc}$ and
$\rho H_{\rm ext}$ terms, where were already accounted for in {\tt
  react\_state}, so our equation becomes
\begin{equation}
\frac{\partial(\rho h)}{\partial t} = -\Ub\cdot\nabla(\rho h) - \rho h\nabla\cdot\Ub 
+ \psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} + \nabla\cdot\kth\nabla T. \label{rhoh equation}
\end{equation}
We define the base state enthalpy evolution equation as
\begin{eqnarray}
\frac{\partial(\rho h)_0}{\partial t} &=& -\nabla\cdot[(\rho h)_0 w_0\eb_r] 
+ \frac{D_0p_0}{Dt} \nonumber \\
&=& -w_0\frac{\partial(\rho h)_0}{\partial r} 
- \underbrace{(\rho h)_0\frac{\partial w_0}{\partial r}+ \psi}_{``(\rho h)_0 ~ \text{force}"}
\enskip .\label{rhoh0 equation}
\end{eqnarray}
Subtracting (\ref{rhoh0 equation}) from (\ref{rhoh equation}) and
rearranging terms gives the perturbational enthalpy equation
\begin{eqnarray}
\frac{\partial(\rho h)'}{\partial t} &=& -\nabla\cdot[(\rho h)'\Ub] 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r} 
+ \nabla\cdot\kth\nabla T\nonumber \\
&=& -\Ub\cdot\nabla(\rho h)' \underbrace{- (\rho h)'\nabla\cdot\Ub 
- \nabla\cdot[(\rho h)_0\Ubt] + (\Ubt \cdot \er)\frac{\partial p_0}{\partial r}
+ \nabla\cdot\kth\nabla T}_{``(\rho h)' ~ \text{force}"}, \label{rhohprime equation}
\end{eqnarray}

Alternately, we can consider an temperature evolution equation, derived
from enthalpy, as:
\begin{equation}
\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] 
+ \nabla \cdot \kth \nabla T
- \sum_k\rho\xi_k\omegadot_k 
+ \rho H_{\rm nuc} + \rho H_{\rm ext}\right\}.
\end{equation}
Again, we neglect the reaction terms, since that will be handled during
the reaction step, so we can write this as:
\begin{equation}
\frac{\partial T}{\partial t} = -\Ub\cdot\nabla T
\underbrace{
+ \frac{1}{\rho c_p}\left\{(1-\rho h_p)\left[\psi
+ (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}\right] 
+ \nabla \cdot \kth \nabla T \right \} }_{``T~\text{force''}} \, .
\label{T equation labeled}
\end{equation}

A final alternative is to consider an evolution equation for $h$
alone.  This can be derived by expanding the derivative of $(\rho h)$
in Eq.~\ref{rhoh equation} and subtracting off $h \times$ the
continuity equation (Eq.~\ref{rho equation}):
\begin{equation}
\frac{\partial h}{\partial t} = -\Ub \cdot \nabla h 
\underbrace{+ \frac{1}{\rho}
\left \{ \psi + (\Ubt \cdot \er )\frac{\partial p_0}{\partial r}
+ \nabla \cdot \kth \nabla T \right \} }_{``h~\text{force''}} \, .
\label{h equation labeled}
\end{equation}

@@

The general procedure for making the $(\rho h)$ edge state is as follows:
\begin{enumerate}
\item predict $(\rho h)'$, $h$, or $T$ to the edges (depending on {\tt
  enthalpy\_pred\_type} ) using {\tt make\_edge\_scal} and the forces
  identified in the appropriate evolution equation
  (Eqs.~\ref{rhohprime equation}, \ref{T equation labeled}, or \ref{h
    equation labeled} respectively).

  The appropriate forces are summaried in table~\ref{table:pred:hforce}.

\item if we predicted $T$, convert this predicted
  edge state to an' intermediate ``enthalpy'' state (the quotes
  indicate that it may be perturbational or full enthalpy) by calling
  the EOS.

\item construct the final enthalpy edge state in {\tt mkflux}.  The
  precise construction depends on what species and enthalpy quantities
  are input to {\tt mkflux}.

\end{enumerate}

\begin{table*}[h]
\centering
\caption{Forcing term into {\tt make\_edge\_scal} \newline}
\label{table:pred:hforce}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{l|c}
\hline
\hline
{\tt enthalpy\_pred\_type} &   {advective force} \\
\hline
1 / {\tt predict\_rhohprime} $((\rho h)^\prime)$ &  $(\Ubt \cdot \er) \frac{\partial p_0}{\partial r} - 
 (\rho h)^\prime \; \nabla \cdot (\Ubt+w_0 \er) - 
 \nabla \cdot (\Ubt (\rho h)_0) + \nabla \cdot \kth \nabla T$ \\
2 / {\tt predict\_h}  $(h)$ & $\frac{1}{\rho} \left [\psi + (\Ubt \cdot \er)
  \frac{\partial p_0}{\partial r} + \nabla \cdot \kth \nabla T \right ]$ \\
3 / {\tt predict\_T\_then\_rhohprime} $(T)$ & $\frac{1}{\rho c_p} \left \{ (1 - \rho h_p) 
   \left [\psi + (\Ubt \cdot \er) \frac{\partial p_0}{\partial r} \right ] + \nabla \cdot \kth \nabla T \right \}$ \\
4 / {\tt predict\_T\_then\_h}  $(T)$ & $\frac{1}{\rho c_p} \left\{ (1 - \rho h_p) \left [\psi + (\Ubt \cdot \er)
\frac{\partial p_0}{\partial r}\right ] +  \nabla \cdot \kth \nabla T \right\}$ \\
\hline
\end{tabular}
\end{table*}

\begin{landscape}
\begin{table}[t]
\caption{Summary of enthalpy edge state construction}
\label{table:pred:hoverview}
\renewcommand{\arraystretch}{1.5}
{\small
\centering
\begin{tabular}{l|l|c|c|c|c}
\hline
\hline
{\tt species\_pred\_type} &   
{\tt enthalpy\_pred\_type} &
{quantity predicted} &
{intermediate} &
{species quantity} &
{final $(\rho h)$} \\[-5pt]
 &
 &
{in {\tt make\_edge\_scal}} &
{``enthalpy'' edge state} &
{available in {\tt mkflux}} &
{edge state} \\
\hline 
1 / {\tt predict\_rhoprime\_and\_X}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'$ & $X$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 2 / {\tt predict\_h} &
  $h$ & $h$ & $X$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'$ & $X$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
1 / {\tt predict\_rhoprime\_and\_X}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h$ & $X$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
\hline
\hline
2 / {\tt predict\_rhoX}  & 1 / {\tt predict\_rhohprime} &
  $(\rho h)'$ & $(\rho h)'$ & $(\rho X)$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
2 / {\tt predict\_rhoX}  & 2 / {\tt predict\_h} &
  $h$ & $h$ & $(\rho X)$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
2 / {\tt predict\_rhoX}  & 3 / {\tt predict\_T\_then\_rhohprime} &
  $T$ & $(\rho h)'$ & $(\rho X)$, $\rho'$ & 
  $\left [ (\rho h)_0 + (\rho h)' \right ]$ \\
2 / {\tt predict\_rhoX}  & 4 / {\tt predict\_T\_then\_h} &
  $T$ & $h$ & $(\rho X)$, $\rho'$ & 
  $\left ( \rho_0 + \rho' \right ) h$ \\
\hline
\end{tabular}
} % end \small
\end{table}
\end{landscape}

@@

\subsection{Computing $\rho h$ at edges}
We use an analogous procedure described in Section \ref{Computing rho
  at edges} for computing $\rho$ at edges to compute $\rho h$ at
edges.  For spherical, however, instead of computing $(\rho h)_0$ to
edges directly, we compute $\rho$ and $h$ separately, and multiply to
get $(\rho h)_0$.

@@

@@

@@

%% \begin{table*}[h]
%% \begin{center}
%% \caption{Quantity that goes into and out of {\tt make\_edge\_scal} \newline}
%% \begin{tabular}{c|c}
%% \hline
%% \hline
%% {h pred type} & {h/T } \\
%% \hline \\[-3mm]
%% 1 & $(\rho h)^\prime$ \\
%% 2 & $h$  \\
%% 3 & $T$  \\
%% 4 & $T$  \\
%% \hline
%% \end{tabular}
%% \end{center}
%% \end{table*}

%% \begin{table*}
%% \begin{center}
%% \caption{Quantity that goes into {\tt mkflux} on edges \newline}
%% \begin{tabular}{c|c|c}
%% \hline
%% \hline
%% {h pred type} & {Species } & {h/T } \\
%% \hline
%% 1 & $X$ and $\rho^\prime$ & $(\rho h)^\prime$ \\
%% 2 & $X$ and $\rho^\prime$ & $h$ \\
%% 3 & $X$ and $\rho^\prime$ & $(\rho h)^\prime$ \\
%% 4 & $X$ and $\rho^\prime$ & $h$ \\
%% \hline
%% \end{tabular}
%% \end{center}
%% \end{table*}
